The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-02 conf_article/neon.tex
Finished the introduction rewrite

In this article, we explore the use of matrix norms for optimizing functions of weight matrices, a crucial problem in training large language models. Moving beyond the spectral norm that underlies the Muon update, we leverage the duals to the Ky Fan k-norms to introduce a \deleted[id=IK, comment={Redundant}]{new} family of Muon-like algorithms we name {\it Fanions}, which happen to be similar to Dion. \replaced[id=IK, comment={This rewrite clarifies the sentence structure, making it clear that the norms (not the algorithms) are being combined.}]{Subsequently, we construct a second family of algorithms, dubbed {\it F-Fanions}, based on the duals of convex combinations of Ky Fan k-norms and the Frobenius norm. One prominent member of this family is {\it F-Muon}.}{Then, we consider their convex combinations with the Frobenius norm and corresponding duals to develop a new family of algorithms we name {\it F-Fanions}, one of them being {\it F-Muon}.} We complement our theoretical analysis with an extensive empirical study of the algorithms across a wide range of tasks and settings, from which it follows that properly-tuned F-Muon is almost on par with Muon, \replaced[id=IK, comment={Alex, clarify this please. It's a very non-obvious and non-clear conclusion, especially when knowing that Muon itself only provides an approximate answer to the lmo.}]{which questions the necessity of finding the exact solution to the problem in linear optimization oracle framework.}

@@

That is what can be said from a practical point of view. \replaced[id=IK, comment={Rewrote this to be more precise. The innovation wasn't just 'using' a norm, but deriving the update from a principled, constrained optimization problem. I think this better reflects the contribution of the Muon paper.}]{From a theoretical perspective, a key innovation of Muon was its principled derivation of the update rule, which emerges as the solution to an optimization problem constrained by the spectral norm \citep{bernstein2025deriving}}{From the perspective of theory, the main innovation of Muon has been an intentional usage of matrix norms, i.e. the spectral norm, to derive the algorithm's update}.

Motivated by the success of Muon, many generalizations and variations of it were proposed. Among the notable ones are Scion \citep{}, Deon \citep{} and Gluon \citep{}. One question still persists however: Why

Based upon recent theoretical advances that explain some theory behind Muon, Scion and Gluon \citep{bernstein2025deriving,kovalev2025understanding,pethick2025training,riabinin2025gluon}, we explore application of other matrix norms to optimization of functions of matrices. As it has been done with Muon, we stipulate that our algorithms' updates be fast to compute.

\replaced[id = IK, comment={Subsequent paragraphs in the introduction are almost completely new.}]{}{}

@@

\section{Problem Statement}
\subsection{Function of a matrix}
    We consider the problem of minimizing a differentiable matrix function $F(\cdot)\colon \Rmn \to \R$
    \begin{equation}\label{eq:mat}
        \min_{\mX \in \Rmn} f(\mX).
    \end{equation}

@@

\section{Experiments}
    \subsection{Randomized Linear Least Squares}
    Since the provided by other authors \citep{kovalev2025understanding,riabinin2025gluon} theoretical guarantees are almost norm-independent, we have to test them in practice.

>>>>

In this article, we explore the use of matrix norms for optimizing functions of weight matrices, a crucial problem in training large language models. Moving beyond the spectral norm that underlies the Muon update, we leverage the duals to the Ky Fan k-norms to introduce a \deleted[id=IK, comment={Redundant}]{new} family of Muon-like algorithms we name {\it Fanions}, which happen to be similar to Dion. \replaced[id=IK, comment={This rewrite clarifies the sentence structure, making it clear that the norms (not the algorithms) are being combined.}]{Subsequently, we construct a second family of algorithms, dubbed {\it F-Fanions}, based on the duals of convex combinations of Ky Fan k-norms and the Frobenius norm. One prominent member of this family is {\it F-Muason}.}{Then, we consider their convex combinations with the Frobenius norm and corresponding duals to develop a new family of algorithms we name {\it F-Fanions}, one of them being {\it F-Muon}.} We complement our theoretical analysis with an extensive empirical study of the algorithms across a wide range of tasks and settings, from which it follows that properly-tuned F-Muon is almost on par with Muon, \replaced[id=IK, comment={Alex, clarify this please. It's a very non-obvious and non-clear conclusion, especially when knowing that Muon itself only provides an approximate answer to the LMO. I provided an alternative phrase.}]{which questions the necessity of finding the exact solution to the problem in linear optimization oracle framework.}{which shows the flexibility in the choice of the norm and opens possibilities for easier-to-compute variants.}

@@

That is what can be said from a practical point of view. \replaced[id=IK, comment={Rewrote this to be more precise. The innovation wasn't just 'using' a norm, but deriving the update from a principled, constrained optimization problem. I think this better reflects the contribution of the Muon paper.}]{From a theoretical perspective, a key innovation of Muon was its principled derivation of the update rule, which emerges as the solution to an optimization problem constrained by the RMS-to-RMS norm (scaled version of spectral norm) \citep{bernstein2025deriving}}{From the perspective of theory, the main innovation of Muon has been an intentional usage of matrix norms, i.e. the spectral norm, to derive the algorithm's update}.

\replaced[id = IK, comment={Subsequent paragraphs in the introduction are very heavily modified.}]{}{}

Motivated by the success of Muon, many generalizations and variations of it were proposed. Among the notable ones are Scion \citep{pethick2025training}, Dion \citep{ahn2025dioncommunicationefficientoptimizerlarge} and Gluon \citep{riabinin2025gluon}. Those researches target Muon's efficiency and establish convergence bounds. One central question, however, remains unanswered:

\textit{In deriving Muon's update step, why constrain by the spectral norm? How would alternative norms affect performance and computational cost?}

In this article, we tackle this question by first showing the connection between matrix norms and corresponding existing algorithms and discuss the theoretic bounds derived for those algorithms. We then leverage the family of norms dual to Ky Fan k-norms to derive a new class of algorithms with low-rank updates, which we call Fanions. Subsequently, we create a second, hybrid family named F-Fanions by constructing convex combinations of the Ky Fan norms with the Frobenius norm and taking dual of that composite norm. Working within the linear minimization oracle (\replaced[id = IK, comment={I think it's better this way.}]{LMO}{lmo}) framework we derive the explicit update formulas for both algorithm families. As it was done with Muon, we stipulate our algorithms to be fast to approximate, for which we utilize the Lanczos algorithm as described in \cref{sec:matrix-side}.

We then compare the performance of the algorithm families across various benchmarks (\cref{sec:experiments}):
\begin{itemize}
    \item Synthetic least squares experiment,
    \item CIFAR-10 airbench,
    \item Pre-training NanoGPT on FineWeb dataset.
\end{itemize}

Our experiments reveal important insights about the role of matrix norms in optimization. On a synthetic least squares problem, we observe a striking discrepancy: while some algorithms converge slowly in terms of loss, they may converge quickly in terms of gradient norm, and vice versa. This suggests that existing theoretical guarantees may not fully explain practical algorithm performance.

Most notably, our experiments on real-world tasks demonstrate that the choice of underlying matrix norm is remarkably flexible. On CIFAR-10 airbench, properly-tuned F-Muon achieves $94.01 \pm 0.14\%$ accuracy, essentially matching Muon's $94.00 \pm 0.13\%$ performance. On NanoGPT pre-training, F-Muon achieves 3.281 cross-entropy loss, only marginally worse than Muon's 3.279. These results show that Muon-like algorithms can maintain competitive performance even when the underlying norm constraint is significantly modified, answering affirmatively the central question posed above. This flexibility suggests potential for developing easier-to-compute variants of successful algorithms like Muon.

@@

\section{Problem Statement}

\subsection{Function of a matrix}
    We consider the problem of minimizing a differentiable matrix function $F(\cdot)\colon \Rmn \to \R$
    \begin{equation}\label{eq:mat}
        \min_{\mX \in \Rmn} f(\mX).
    \end{equation}

@@

\section{Experiments}\label{sec:experiments}
    \subsection{Randomized Linear Least Squares}
    Since the provided by other authors \citep{kovalev2025understanding,riabinin2025gluon} theoretical guarantees are almost norm-independent, we have to test them in practice.
